Title Goes Here

### 2021-04-06 analysis/paper/paper.Rmd
Muchos cambios

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 42,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

lab_lev <- AddSuffix(" hPa")
lab_cplx <- c(R = "Real", 
              I = "Imaginary")
lab_PC <- c(PC1 = "EOF1", 
            PC2 = "EOF2")

@@

```{r compute-eof-naive}
eof_naive <- era5[lev %in% c(50, 200) & lat %between% c(-90, -20)] %>% 
  # .[, hgt := Anomaly(hgt), by = .(season(time), lon, lat, lev)] %>%
  .[, hgt  := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(time, lev, lat)] %>% 
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "EOF"))), 
    by = lev]

explained <- eof_naive[, eof[[1]]$sdev, by = lev] %>% 
  .[order(lev, EOF)] %>% 
  .[, scales::percent(r2, accuracy = .1)]

@@

```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=3.7, fig.height=2.2}
eof_naive[, eof[[1]]$right, by = lev] %>%
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = hgt, fill = ..level..),
                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0), 
                    range = c(0.01, 0.2)) +

  geom_qmap(~.x[lat <= -20], size = 0.1) +
  scale_fill_divergent_discretised(guide = "none")  +
  scale_y_latitude() +
  scale_x_longitude() +
  coord_polar() +
  theme(axis.text = element_blank()) +
  facet_grid(lev~EOF, labeller = labeller(lev = lab_lev))  +
  tagger::tag_facets("rc")  +
  theme(panel.spacing = grid::unit(-.7, "lines"))
```

In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase \textcolor{red}{(Agregaría algunas referencias)}. For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous September through October \textcolor{red}{(November supongo. Sería bueno que ya arriba indiques lo de SON y ya después lo menciones directamente así)} geopotential height at 50 hPa and 200 hPa. It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3  \textcolor{red}{(No es visualmente obvia una estructura de onda 3. Solo se ven 4 centros)} pattern shifted by 1/4 wavelength. Similarly, EOFs 2 and 3 at 200 hPa represent the a zonal wave 3 pattern shifted by 1/4 wavelength. Since these pairs of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

@@

```{r corr-ceof-splitted}
ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
  copy() %>%
  .[, mag := abs(hgt)] %>%
  .[, item := paste(lev, PC, sep = "_")] %>%
  widyr::pairwise_cor(item, time, mag) %>%
  setDT() %>%
  # .[]
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
  .[tstrsplit(item1, "_")[[1]] != 50] %>%
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  .[, r2 := round(correlation^2, 2)] %>%
  dcast(item1 ~ item2, value.var = "r2") %>%
  setnames("item1", "200 hPa") %>%
  kbl2(booktabs = TRUE, caption = "Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.") %>%
  column_spec2(2:4, background = spec_color2(.table[[.col]], option = background_palette,
                                            scale_from = c(0, 1)),
               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
  add_header_above(c("", "50 hPa" = 3)) %>%
  kable_classic() 
```

Another characteristic seen in Figure\ \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EOF1 at 50 hPa  \textcolor{red}{(En la figura que yo veo, no diría que no similares. Aclara mejor a qué te referis)}. The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa, excecpt for the behaviour above the Indian Ocean. Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF across levels. There's a high degree of correlation between the respective EOF1 and EOF2 at each level. Both the spatial similarities and temporal correlations suggest some level of joint variability between levels. \textcolor{red}{(Revisar el parráfo considerando el comentario)}

Both observations \textcolor{red}{(¿Qué serían las dos osbervaciones?)}  motivate the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

@@

# rotations_pc1 <- lapply(angles, function(a) {
#   ceof$eof[[1]]$left[PC == "PC1"] %>%
#     copy() %>%
#     .[, hgt := rotate(hgt, a)] %>%
#     sep_ReIm() %>%
#     .[, .(mean = mean(hgt)), by = part] %>%
#     .[, rotation := a]
# }) %>%
#   rbindlist()
# 
# 
# best_rotation_pc1 <- rotations_pc1[part == "Real"][which.min(mean)]$rotation

@@

with_o3 <- ceof$eof[[1]]$left %>% 
  .[PC == "PC1"] %>% 
  .[wave1_o3, on = "time"]

@@

best_rotation_pc1 <- rotations_o3[part == "Real"][which.max(cor)]$rotation

@@

rotations_pc2 <- lapply(angles, function(a) {
  with_enso %>%
    .[, hgt2 := rotate(hgt, a)] %>%
    .[, .(R = cor(Re(hgt2), oni),
          I = cor(Im(hgt2), oni))] %>%
    .[, rotation := a]
}) %>%
  rbindlist()

best_rotation_pc2 <- rotations_pc2[I > 0][which.min(abs(R))]$rotation

eof_rotated_left <- ceof$eof[[1]]$left %>%
  copy() %>%
  .[PC == "PC2", hgt := rotate(hgt, best_rotation_pc2)] %>%
  .[PC == "PC1", hgt := rotate(hgt, best_rotation_pc1)]

eof_rotated_right <-  ceof$eof[[1]]$right %>%
  copy() %>%
  .[PC == "PC2", hgt := rotate(hgt, best_rotation_pc2)] %>%
  .[PC == "PC1", hgt := rotate(hgt, best_rotation_pc1)]

@@

```{r rotations, fig.cap = "(ref:rotations-cap)", fig.width=width_column, fig.height=2.5}
r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
    paste("2")
}, "", ""))

rotations_o3 %>%
  ggplot(aes(rotation, cor)) +
  geom_vline(xintercept = best_rotation_pc1) +
  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
  geom_line(aes(color = part)) +
  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 90*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous("Correlation") +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +

rotations_pc2 %>%
  melt(id.vars = "rotation") %>%
  ggplot(aes(rotation, value^2)) +
  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
  geom_vline(xintercept = best_rotation_pc2) +
  geom_line(aes(color = variable)) +
  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 90*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous(r2) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +

  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```

\textcolor{red}{(No mostraría la figura 2. Me parece que agrega una complejidad extra y no sé entiende demasiado, sino se la explica muy bien)}

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`. This also nearly minimises the correlation with the imaginary part (Figure\ \@ref(fig:rotations)a).

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part (Figure\ \@ref(fig:rotations)b).

@@

## Zonal anomalies circulation modes (no me convence del todo, pero debería ser algo así)

```{r define-plot_ceof}
plot_ceof <- function(data, n = 1:2, formula = lev  ~  PC) {

@@

  var <- data[, cut(eof[[1]], n)$sdev, by = .(season)] %>%
    .[, setNames(paste0(PC, " (", scales::percent(r2), ")"),
                 PC)]

@@

    geom_qmap(~ .x[lat <= -20]) +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-"),
                          guide = "none") +
    facet_grid(formula,
               labeller = labeller(lev = lab_lev,
                                   PC = var)) +
    coord_quickmap(ylim = c(NA, -20)) +
    theme(legend.title = element_blank())
}
```

```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019. Real part in shading, imaginary part in contours", fig.width=width_column, fig.height=width_column}
ceof[season == "SON"] %>%
  plot_ceof() +
  coord_polar() +
  axis_labs_smol +
  tag_facets(tag = "cr")
```

```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2). The first mode (CEOF1) explains 82% of the variance, while the second mode (CEOF1) explains a smaller fraction (7%).  In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the complex temporal EOF. The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive): The first is mainly a wave 1, while the second is a wave 3.

The temporal variability for both components of the cEOF1 have non zero mean (Fig.\ \@ref(fig:ceofs-2)). This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean. As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field.

\textcolor{red}{(Acá creo que habría que ampliar un poco más la discusión. diciendo en más detalle lo que se ve en la parte espacial (donde se ven los centros más importantes, qué diferencias hay entre estratósfera y troposféra) y temporal, por ej. qué tipo de variabilidad se observa, si están correlacionadas)}

@@

era_more <- rbind( 
  ReadNetCDF(ERA5_BE(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50))),
  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50)))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
  .[, season := season(time)] %>%
  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
  .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, PC, time)]

@@

```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=3, fig.width = 7, fig.env = "figure*"}
# Check that the extended index coincides with the
# original index.
cors <- era_more %>%
  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "PC", "part")] %>%
  na.omit() %>%
  .[, cor(hgt, i.hgt), by = .(PC, part)]

@@

trends <- era_more %>%
  .[, FitLm(hgt, time, se = TRUE), by = .(part, PC)] %>%
  rm_intercept() %>%
  .[, p_val := Pvaluate(estimate, std.error, df)] %>%
  .[, p_report := report_p(p_val)]

# era_less <-   ceof %>%
#   .[, eof[[1]]$left, by = season] %>%
#   sep_ReIm(hgt) %>%
#   .[season == "SON"] %>% 
#   .[, hgt := as.numeric(scale(hgt)), by = .(part, PC)]

era_more %>%
  copy() %>% 
  .[, ref := year(time) > 1979] %>% 
   # .[, hgt := scale(hgt, center = FALSE, scale = sd(hgt[ref])), by = .(part, PC)] %>%
  ggplot(aes(time, hgt*1e3)) +
  # Zero line + rangeframe
  geom_hline(data = ~.x[, mean(hgt)*1e3, by  = .(PC, part)], 
             aes(yintercept = V1), size = 0.1, color = "gray30") +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  geom_line() +
  # geom_line(data = era_less) +
  scale_alpha_manual(values = c(1, 0.2)) +
  geom_smooth(method = "lm", color = "black") +
  scale_x_datetime(NULL, date_labels = "%Y") +
  scale_y_continuous("EOF") +

  facet_grid(PC ~ part, labeller = labeller(PC = lab_PC), scales = "free_y") +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  tag_facets("cr") +
  panel_background
```

Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950. \textcolor{red}{(Me viene una duda metodológica que vendría bien aclarar. Cuando decis que lo extendes, ¿es que lo calculaste para todo ese período o que proyectas el modo en los otros tiempos?)} There is a significant upward trend in the real component of cEOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex components of cEOF2. The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown). \textcolor{red}{(¿Qué interpretación se le podría dar a esto? ¿Que las amplitudes de las anomalías asociadas al modo son cada vez mayores?)}

## Regressions  (me parece que habría que decirlo de una forma más precisa o juntarlo a la sección anterior)

@@

The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series. To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields. 
```{r compute-regression}
regr <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))),
    by = .(season, PC)] %>%
  dcast(time ~ PC, value.var = c("Real", "Imaginary")) %>%
  .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
  .[lev %in% c(850, 700, 200, 50)] %>%
  # .[variable == "hgt"] %>%
  .[, FitLm(value, Imaginary_PC1, Real_PC1, Imaginary_PC2, Real_PC2, se = TRUE),
    by = .(lev, lon, lat, variable)] %>%
  .[, c("term", "PC") := tstrsplit(term, "_")] %>%
  rm_intercept() %>%
  .[, term := factor_ReIm(term)]
```

>>>>

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 41,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

lab_lev <- AddSuffix(" hPa")
lab_cplx <- c(R = "Real", 
              I = "Imaginary")

@@

```{r compute-eof-naive}
eof_naive <- era5[lev %in% c(50, 200) & lat %between% c(-90, -20)] %>% 
  # .[, hgt := Anomaly(hgt), by = .(season(time), lon, lat, lev)] %>%
  .[, hgt  := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(time, lev, lat)] %>% 
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "cEOF"))), 
    by = lev]

explained <- eof_naive[, eof[[1]]$sdev, by = lev] %>% 
  .[order(lev, cEOF)] %>% 
  .[, scales::percent(r2, accuracy = .1)]

@@

```{r eof-naive, fig.cap = "(ref:eof-naive-cap)", fig.width=2.5, fig.height=3}
eof_naive[, eof[[1]]$right, by = lev] %>%
  .[lev == 50] %>% 
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = hgt, fill = ..level..),
                    breaks = AnchorBreaks(0, 0.01, exclude = 0)) +
  geom_contour_tanaka(aes(z = hgt), breaks = AnchorBreaks(0, 0.01, exclude = 0), 
                    range = c(0.01, 0.2)) +

  geom_qmap(~.x[lat <= -20], size = 0.1) +
  scale_fill_divergent_discretised(guide = "none")  +
  scale_y_latitude() +
  scale_x_longitude() +
  coord_polar() +
  theme(axis.text = element_blank()) +
  facet_wrap(.~cEOF, ncol = 2, labeller = labeller(lev = lab_lev))  +
  tagger::tag_facets("rc")  +
  theme(panel.spacing = grid::unit(-.7, "lines"))
```

In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase [@horel1984]. For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous SON geopotential height at 50 hPa and 200 hPa. It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave pattern with shorter wavenumber shifted by 1/4 wavelength. A similar EOF structure can be seen in 200 hPa (not shown). Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

@@

```{r corr-ceof-splitted}
ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
  copy() %>%
  .[, mag  := abs(hgt)] %>%
  .[, item := paste(lev, cEOF, sep = "_")] %>%
  widyr::pairwise_cor(item, time, mag) %>%
  setDT() %>%
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
  .[tstrsplit(item1, "_")[[1]] != 50] %>%
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  .[, r2 := round(correlation^2, 2)] %>%
  dcast(item1 ~ item2, value.var = "r2") %>%
  setnames("item1", "200 hPa") %>%
  kbl2(booktabs = TRUE, caption = "Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.") %>%
  column_spec2(2:4, background = spec_color2(.table[[.col]], option = background_palette,
                                            scale_from = c(0, 1)),
               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
  add_header_above(c("", "50 hPa" = 3)) %>%
  kable_classic() 
```

Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels. There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level. The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown)

Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability. This motivates the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

@@

# rotations_cEOF1 <- lapply(angles, function(a) {
#   ceof$eof[[1]]$left[cEOF == "cEOF1"] %>%
#     copy() %>%
#     .[, hgt := rotate(hgt, a)] %>%
#     sep_ReIm() %>%
#     .[, .(mean = mean(hgt)), by = part] %>%
#     .[, rotation := a]
# }) %>%
#   rbindlist()
# 
# 
# best_rotation_cEOF1 <- rotations_cEOF1[part == "Real"][which.min(mean)]$rotation

@@

with_o3 <- ceof$eof[[1]]$left %>% 
  .[cEOF == "cEOF1"] %>% 
  .[wave1_o3, on = "time"]

@@

best_rotation_cEOF1 <- rotations_o3[part == "Real"][which.max(cor)]$rotation

@@

rotations_cEOF2 <- lapply(angles, function(a) {
  with_enso %>%
    .[, hgt2 := rotate(hgt, a)] %>%
    .[, .(R = cor(Re(hgt2), oni),
          I = cor(Im(hgt2), oni))] %>%
    .[, rotation := a]
}) %>%
  rbindlist()

best_rotation_cEOF2 <- rotations_cEOF2[I > 0][which.min(abs(R))]$rotation

eof_rotated_left <- ceof$eof[[1]]$left %>%
  copy() %>%
  .[cEOF == "cEOF2", hgt := rotate(hgt, best_rotation_cEOF2)] %>%
  .[cEOF == "cEOF1", hgt := rotate(hgt, best_rotation_cEOF1)]

eof_rotated_right <-  ceof$eof[[1]]$right %>%
  copy() %>%
  .[cEOF == "cEOF2", hgt := rotate(hgt, best_rotation_cEOF2)] %>%
  .[cEOF == "cEOF1", hgt := rotate(hgt, best_rotation_cEOF1)]

@@

```{r rotations, fig.cap="(ref:rotations-cap)", fig.height=2.5, fig.width=width_column, include=FALSE}
r2 <- expression(`$r^2$` = paste("", "r", phantom()^{
    paste("2")
}, "", ""))

rotations_o3 %>%
  ggplot(aes(rotation, cor)) +
  geom_vline(xintercept = best_rotation_cEOF1) +
  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
  geom_line(aes(color = part)) +
  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 90*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous("Correlation") +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +

rotations_cEOF2 %>%
  melt(id.vars = "rotation") %>%
  ggplot(aes(rotation, value^2)) +
  geom_hline(yintercept = 0,  size = 0.2, color = "gray50") +
  geom_vline(xintercept = best_rotation_cEOF2) +
  geom_line(aes(color = variable)) +
  scale_x_continuous("Rotation", breaks = seq(-pi, pi, by = 90*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous(r2) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_cplx) +

  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```
For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`. This also nearly minimises the correlation with the imaginary part.

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.

@@

## cEOF patterns

```{r define-plot_ceof}
plot_ceof <- function(data, n = 1:2, formula = lev  ~  cEOF) {

@@

  var <- data[, cut(eof[[1]], n)$sdev, by = .(season)] %>%
    .[, setNames(paste0(cEOF, " (", scales::percent(r2), ")"),
                 cEOF)]

@@

    geom_qmap(~ .x[lat <= -20]) +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-"),
                          guide = "none") +
    facet_grid(formula,
               labeller = labeller(lev = lab_lev,
                                   cEOF = var)) +
    coord_quickmap(ylim = c(NA, -20)) +
    theme(legend.title = element_blank())
}
```

```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019. Real part in shading, imaginary part in contours.", fig.width=width_column, fig.height=width_column}
ceof[season == "SON"] %>%
  plot_ceof() +
  coord_polar() +
  axis_labs_smol +
  tag_facets(tag = "cr")
```

```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, cEOF)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous(NULL) +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(cEOF ~ ., switch = "y") +
  theme(panel.spacing = grid::unit(1.5, "lines"), strip.placement = "outside")  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2). The first mode (CEOF1) explains 82% of the variance, while the second mode (CEOF1) explains a smaller fraction (7%).  In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the complex temporal EOF. The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive).

The first cEOF (Fig.\ \@ref(fig:eof-naive) column\ a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes. At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is locates at around 175ºE indicating a westerly shift in phase.

The second cEOF (Fig.\ \@ref(fig:eof-naive) column\ b) show also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales. In particular, it's dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector. This modulated amplitude is especially apparent at 200 hPa. There is no apparent phase shift, suggesting a rather barotropic structure.

The temporal variability for both components of the cEOF1 have non zero mean (Fig.\ \@ref(fig:ceofs-2)). This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean. As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field. There are no significant simultanous correlation between the time series.

@@

era_more <- rbind( 
  ReadNetCDF(ERA5_BE(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50))),
  ReadNetCDF(ERA5(), vars = c(hgt = "z"),
             subset = list(latitude = range(pattern$lat),
                           level = list(200, 50)))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(hgt = mean(hgt)), by = .(lev, lon, lat, time = seasonally(time))] %>%
  .[, season := season(time)] %>%
  .[, hgt := hgt/sd(hgt), by = .(lev)] %>%
  .[pattern, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, .(hgt = weighted.mean(hgt*EOF, w = cos(lat*pi/180))),  by = .(part, cEOF, time)]

@@

```{r extended-series, fig.cap = "(ref:extended-series-cap)", fig.height=3, fig.width = 7, fig.env = "figure*"}
# Check that the extended index coincides with the
# original index.
cors <- era_more %>%
  .[sep_ReIm(ceof$eof[[1]]$left), on = c("time", "cEOF", "part")] %>%
  na.omit() %>%
  .[, cor(hgt, i.hgt), by = .(cEOF, part)]

@@

trends <- era_more %>%
  .[, FitLm(hgt, time, se = TRUE), by = .(part, cEOF)] %>%
  rm_intercept() %>%
  .[, p_val := Pvaluate(estimate, std.error, df)] %>%
  .[, p_report := report_p(p_val)]

# era_less <-   ceof %>%
#   .[, eof[[1]]$left, by = season] %>%
#   sep_ReIm(hgt) %>%
#   .[season == "SON"] %>% 
#   .[, hgt := as.numeric(scale(hgt)), by = .(part, cEOF)]

era_more %>%
  copy() %>% 
  .[, ref := year(time) > 1979] %>% 
   # .[, hgt := scale(hgt, center = FALSE, scale = sd(hgt[ref])), by = .(part, cEOF)] %>%
  ggplot(aes(time, hgt*1e3)) +
  # Zero line + rangeframe
  geom_hline(data = ~.x[, mean(hgt)*1e3, by  = .(cEOF, part)], 
             aes(yintercept = V1), size = 0.1, color = "gray30") +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  geom_line() +
  # geom_line(data = era_less) +
  scale_alpha_manual(values = c(1, 0.2)) +
  geom_smooth(method = "lm", color = "black") +
  scale_x_datetime(NULL, date_labels = "%Y") +
  scale_y_continuous("EOF") +

  facet_grid(cEOF ~ part, scales = "free_y") +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  tag_facets("cr") +
  panel_background
```

Both cEOF shot year-to-year variability but show no evidence of decadal variabilty. The relatively short timespan (`r diff(range(unique(year(era5$time))))` years) hampers our habilty to detect longer trends. Therefore, we extend the cEOF time series using the preliminary ERA5 back extension going back to 1950 by projecting the older geopogential height zonal anomaly fields onto the spatial patterns shown in Figure\ \@ref(fig:ceofs-1).

The extended cEOF timeseires are shown in Figure\ \@ref(fig:extended-series). There is a significant upward trend in the real component of cEOF1 (Fig.\ \@ref(fig:extended-series)a.1, p-value\ `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2. The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).

This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1. A similar observation was made by @raphael2003, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies.

@@

The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series. To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.
